\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-02-28 fluB.tex
Change subsection headers.
New subsection headers for merged paragraphs.

\subsection*{Maintenance of diversity and co-assortment of PB1, PB2 and HA gene segments}
The differentiation into Vic and Yam lineages can be seen in all segments (Figure \ref{genomeGrid}).
Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
On some segments, either the Victoria or Yamagata lineage has become fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment, seen as modern viruses deriving completely from either Victoria or Yamagata lineage (yellow vs purple bars in Figure \ref{genomeGrid}).
This pattern is apparent in the PA, NP, NA, MP and NS segments.
However, the PB1, PB2 and HA segments of modern viruses are derived from both Victoria and Yamagata lineages.
Consistent with fixation of Victoria or Yamagata lineages, the PA, NP, NA, MP and NS segments periodically lose diversity, while maintenance of parallel Victoria and Yamagata lineages results in continually increasing diversity in segments PB1, PB2 and HA (Figure \ref{tmrcaOT}).
The PB1, PB2 and HA segments from present-day viruses maintain a common ancestor in $\sim$1983 and thus accumulate genetic diversity since the split of those segments into Vic and Yam lineages, while other segments often lose diversity with ancestors to present-day viruses appearing between $\sim$1991 and $\sim$1999.

@@

\subsection*{Lineage composition of circulating PB1-PB2-HA constellations}
Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) since 1997 all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations (Figure \ref{stateTime}).

@@

\subsection*{PB1, PB2 and HA trees experience reassortment `distance' effects}
SPR (subtree prune and regraft) distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between the two trees and is a measure of tree-to-tree similarity.
Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization (see Methods).
If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to observe a lower reassortment rate between them, which would manifest as small-scale similarities between phylogenetic trees of those segments.
In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the raw (\textit{i.e} both within and between lineages) reassortment rate.

@@

\subsection*{Lineage-specific substitutions in PB1, PB2 and HA segments}
Linkage disequilibrium (LD) is a measure of association between alleles at different polymorphic loci within a population.
To estimate LD no other information, except for alleles at polymorphic amino acid sites in the dataset, is required.
Although not an independent validation of association between PB1, PB2 and HA segments, linkage disequilibrium analyses show that Victoria and Yamagata lineages of these segments have accumulated lineage-specific amino acid substitutions.
In Figure \ref{segmentLD} the signature of Vic and Yam lineage-specific substitutions is greater mean $\chiSq$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
In some cases this value exceeds the estimated mean LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA have higher mean LD than mean LD when PB1 is compared against itself).

>>>>

\subsection*{Analysis of reassortment patterns across Victoria and Yamagata lineages}
The differentiation into Vic and Yam lineages can be seen in all segments (Figure \ref{genomeGrid}).
Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
On some segments, either the Victoria or Yamagata lineage has become fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment, seen as modern viruses deriving completely from either Victoria or Yamagata lineage (yellow vs purple bars in Figure \ref{genomeGrid}).
This pattern is apparent in the PA, NP, NA, MP and NS segments.
However, the PB1, PB2 and HA segments of modern viruses are derived from both Victoria and Yamagata lineages.
Consistent with fixation of Victoria or Yamagata lineages, the PA, NP, NA, MP and NS segments periodically lose diversity, while maintenance of parallel Victoria and Yamagata lineages results in continually increasing diversity in segments PB1, PB2 and HA (Figure \ref{tmrcaOT}).
The PB1, PB2 and HA segments from present-day viruses maintain a common ancestor in $\sim$1983 and thus accumulate genetic diversity since the split of those segments into Vic and Yam lineages, while other segments often lose diversity with ancestors to present-day viruses appearing between $\sim$1991 and $\sim$1999.

@@

Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) since 1997 all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations (Figure \ref{stateTime}).

@@

\subsection*{Analysis of within-lineage reassortment patterns}
SPR (subtree prune and regraft) distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between the two trees and is a measure of tree-to-tree similarity.
Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization (see Methods).
If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to observe a lower reassortment rate between them, which would manifest as small-scale similarities between phylogenetic trees of those segments.
In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the raw (\textit{i.e} both within and between lineages) reassortment rate.

@@

Linkage disequilibrium (LD) is a measure of association between alleles at different polymorphic loci within a population.
To estimate LD no other information, except for alleles at polymorphic amino acid sites in the dataset, is required.
Although not an independent validation of association between PB1, PB2 and HA segments, linkage disequilibrium analyses show that Victoria and Yamagata lineages of these segments have accumulated lineage-specific amino acid substitutions.
In Figure \ref{segmentLD} the signature of Vic and Yam lineage-specific substitutions is greater mean $\chiSq$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
In some cases this value exceeds the estimated mean LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA have higher mean LD than mean LD when PB1 is compared against itself).
